Species turnover reveals hidden effects of decreasing Nitrogen deposition in mountain hay meadows

### 2018-06-18 Manuscript.Rmd
Einleitung ergänzt, Referenzen eingefügt.

---
title: Species turnover reveal hidden effects of decreasing Nitrogen deposition in mountain hay meadows
preprint: false
author: 
  - name: Tobias Roth
    affiliation: 1, 2
    corresponding: true
    email: t.roth@unibas.ch
  - name: Lukas Kohli
    affiliation: 2
  - name: Beat Rihm
    affiliation: 3
  - name: Reto Meier
    affiliation: 4    
affiliation:
  - code: 1
    address: Zoological Institute, University of Basel, Basel, Switzerland
  - code: 2
    address: Hintermann  Weber AG, Austrasse 2a, 4153 Reinach, Switzerland
  - code: 3
    address: Meteotest,Fabrikstrasse 14, 3012 Bern, Switzerland 
  - code: 4
    address: Federal Office for the Environment (FOEN), Air Pollution Control and Chemicals Division, 3003 Bern, Switzerland
abstract: >
  Nitrogen (N) deposition is a major threat to biodiversity of many habitats. The recent introduction of cleaner technologies in Switzerland has let to reductions in the emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition. We infered different drivers of community change (i.e. Nitrogen deposition, climate warming, land-use change) in Swiss mountain hay meadows using data from the Swiss biodiversity monitoring.

@@

# Introduction {-}
Nitrogen deposition is the entry of reactive nitrogen compounds into water and soil via deposition through the air. Toghether with land-use and climate change, Nitrogen deposition is one of the major threats to biodiversity [@Sala2000; @Bobbink2010]. Since Nitrogen is an essential plant nutrient and many species-rich communities are adapted to conditions of low N availability [@Vitousek1997], the addition of Nitrogen strongly alter these communities [@@Sala2000]. While there is strong evidence for the reduction in diversity of species-rich grasslands in Europe due to increased Nitrogen deposition [@Stevens2006; @Maskell2010; @Dupre2010], mountain grasslands have received less attention [@Humbert2016].

Since the late 1980s measures to reduce atmospheric pollution have successfully reduced emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition in many parts of Europe [@Torseth2012](Maas and Grennfelt, 2016). However, it is an open question whether and how fast the reduction in N deposition rates will lead to the recovery of extant plant communities.

Quantifying the potential recovery of biodiversity in response to reducing air pollution requires an alternative to the space-for-time approach, ideally monitoring long-term community dynamics on permanent plots (Duprã, C. et al. Changes in species richness and composition in European acidic grasslands over the past 70 years: the contribution of cumulative atmospheric nitrogen deposition. Glob. Change Biol. 16, 344–357, (2010).

One useful approach to understanding biodiversity change is through estimates of biodiversity turnover reflecting both immigration and extinction, often in a closed range of values (Hillebrand et al-2018.pdf).

Observation data of vascular plants from field recordings shows negative correlations of Nitrogen deposition with species richness and species composition in mountain hay meadows (Roth et al. 2013).

In Switzerland, grassland accounts for 70 per cent of the agricultural land. With extensive cultivation, permanent grassland has a very high biodiversity. This applies in particular to the meadows of the Swiss mountain region (Schlup et al. 2013). In the alpine region, meadows with a high plant diversity are of great importance agronomically (Leiber et al. 2005), for the image of the products and for the conservation of biodiversity (Hohl, 2006).

NOx emissions in Switzerland decreased by 46% between 1990 and 2010 and NH3 emissions by 14% (Maas and Grennfelt, 2016)

The meta-analysis of Humbert et al. (2016) showed management and abiotic factors to interact with nitrogen addition to drive plant community changes: The overall effects of N addition on plant species richness and diversity (Shannon index) were more pronounced in managed experimental sites with warmer summers. Management regimes of accessible mountainous areas are intensified, while poorly accessible mountainous areas are being abandoned (Tasser and Tappeiner, 2002; Graf et al., 2014, Strebel and Bühler, 2015). This change in traditional management regimes is considered as a major driver that endangers grassland biodiversity in mountainous regions of Europe (Niedrist et al., 2009; Homburger and Hofer, 2012).

On European mountain summits, plants shifted upwards in the last decades (Rumpf et al. 2018). The authors conclude that «These results suggest that recent climate warming interacted with airborne nitrogen deposition in driving the observed dynamics. As changes increased more for species from lower elevations and eutrophic plant species showed the strongest increase in abundance, these findings might be relevant for mountain hay meadows, too.

>>>>

---
title: Species turnover reveals hidden effects of decreasing Nitrogen deposition in mountain hay meadows
preprint: false
author: 
  - name: Tobias Roth
    affiliation: 1, 2
    corresponding: true
    email: t.roth@unibas.ch
  - name: Lukas Kohli
    affiliation: 2
  - name: Beat Rihm
    affiliation: 3
  - name: Reto Meier
    affiliation: 4    
  - name: Valentin Amrhein
    affiliation: 1    
affiliation:
  - code: 1
    address: Zoological Institute, University of Basel, Basel, Switzerland
  - code: 2
    address: Hintermann  Weber AG, Austrasse 2a, 4153 Reinach, Switzerland
  - code: 3
    address: Meteotest,Fabrikstrasse 14, 3012 Bern, Switzerland 
  - code: 4
    address: Federal Office for the Environment (FOEN), Air Pollution Control and Chemicals Division, 3003 Bern, Switzerland
abstract: >
  Nitrogen (N) deposition is a major threat to biodiversity of many habitats. The recent introduction of cleaner technologies in Switzerland has let to reductions in the emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition. We infered different drivers of community change (i.e. Nitrogen deposition, climate warming, land-use change) in Swiss mountain hay meadows using data from the Swiss biodiversity monitoring.

@@

# Introduction {-}
Nitrogen deposition is the entry of reactive nitrogen compounds into water and soil via deposition through the air. Toghether with land-use and climate change, Nitrogen deposition is one of the major threats to biodiversity [@Sala2000; @Bobbink2010]. Since Nitrogen is an essential plant nutrient and many species-rich communities are adapted to conditions of low N availability [@Vitousek1997], the addition of Nitrogen strongly alter these communities [@Sala2000]. While there is strong evidence for the reduction in diversity of species-rich grasslands in Europe due to increased Nitrogen deposition [@Stevens2006; @Maskell2010; @Dupre2010], mountain grasslands have received less attention [@Humbert2016].

Since the late 1980s measures to reduce atmospheric pollution have successfully reduced emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition in many parts of Europe [@Torseth2012; @Fowler2007]. However, it is an open question whether and how fast the reduction in N deposition rates will lead to the recovery of extant plant communities (>>REF EINFÜGEN<<).

>> Kapitel zu konstante Communities und viel Turnover einfügen und dann erklären, dass wir eine Idee von Community assembly verwendet haben.

One useful approach to understanding biodiversity change is through estimates of biodiversity turnover reflecting both colonization and local extinction [@Hillebrand2018].

In Switzerland, grassland accounts for 70 per cent of the agricultural land. With extensive cultivation, permanent grassland has a very high biodiversity. This applies in particular to the meadows in the alpine region, where meadows with a high plant diversity are of great importance agronomically (Leiber et al. 2005), for the image of the products and for the conservation of biodiversity (Hohl, 2006). Observation data of vascular plants from field recordings shows negative correlations of Nitrogen deposition with species richness and species composition in mountain hay meadows [@Roth2013]. However, between 1990 and 2010 NOx emissions in Switzerland decreased by 46% and NH3 emissions by 14% (>>REF EINFÜGEN: Maas and Grennfelt, 2016<<), suggesting that mountain hay meadows should have recoverd.

However, mountain regions are threatened also by other global change drivers. In Switzerland, temperatures increased from 1959 to 2008 at all altitudes, with an average warming rate of 0.35&degC per decade, which is about 1.6 times the northern hemispheric warm- ing rate (Ceppi et al. 2010). In autumn and early winter, the warming rate depended on altitude with strongest trends at low altitudes, whereas in spring slightly stronger trends are found at altitudes close to the snow line. This climate warming is likely to directly affect or interact with Nitrogen deposition to drive plant community changes [@Humbert2016]. At the landscape scale, changes of plant communities were found that could be attributed to climate warming [@Roth2014]. Particularly, the shift of plant communities at mountain summits seems to be the restult of recent climate warming interacting with airborne nitrogen deposition [@Rumpf2018]. Furthermore, traditional management regimes are currently changing wich also has major impacts on grassland biodiversity in mountainous regions of Europe [@Niedrist2009; @Homburger2012]. In Switzerland  Management regimes of accessible mountainous areas are intensified, while poorly accessible mountainous areas are being abandoned [@Tasser2002; @Strebel2015].

Here, we used the data from the Swiss biodiversity monitoring [@Weber2004] and addressed the following questions:

>>1.  community measures , 2. Compare to turnover and 3. compare incicator values of species that newly colonized or disappeard (local extinction) from sites with random expectation.
